Estimating length of SARS-CoV-2 from doubly interval censored data with missed infections

### 2024-03-19 transmission.tex
spelling / grammar check

In this chapter, I use the estimate of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
I will show that the CIS data is sufficient; demonstrating the value of surveys of this type for pandemic surveillance.

@@

First, in \cref{backcalc}, I use a back-calculation approach, based on \cref{E-inc-prev:eq:EPt}.
This approach enables estimation of incidence within (sub)populations of England.
There are several attractive features about this approach in the context here.
The backcalculation approach makes no assumptions about the transmission process, which have the potential to introduce bias.
It can also easily incorporate arbitrary covariates to adjust for non-response bias.
The scale of the CIS and desire to correct for non-response bias presents challenges for this methodology, especially within the limited computation of the SRS environment.

@@

The first approach, which I call the \emph{incidence-inferred} approach, is a single statistical model where incidence is inferred, with prevalence as a derived quantity.
The basis of the statistical model, in a Bayesian framework, would be as follows.
\begin{align}
    \vec{Z} \mid \vec{\phi} &\dist g_Z(\vec{\phi}) \\
    P_{i,t} &= \matr{S} \vec{Z}_i \\
    Y_{i,t} \mid \vec{P} &\dist \BB(n_{i,t}, P_{i,t})
\end{align}
where $\vec{Z}$ is a vector of incidence proportion for each strata and day defined by some model $g$ with parameters $\vec{\phi}$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\matr{S}$ is the survival matrix defined in \cref{E-inc-prev:eq:Smatrix}; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
The model $g_Z$ would be specified to share information across time and/or strata.
This both helps stabilise the estimation (see\todo{ref wherever I end up explaining backcalculation stability issues}) and allows borrowing of information across strata.
Borrowing information is required because some strata and time combinations have few positives (fewer than five per day).
The relationship between $P$ and $Z$ is from \cref{E-inc-prev:eq:EPt-matrix}, although dividing both sides by $\Npop$ to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.

The second approach, which I call the \emph{prevalence-inferred} approach takes a two-step approach.
First, it uses the statistical model from \cref{E-biology-data:sec:MRP} to estimate prevalence.
Then, it transforms the posterior distribution of prevalence to a posterior distribution of incidence using the forward substitution as in \cref{E-inc-prev:eq:forward-substitute}.
At a high-level, the statistical model is:
\begin{align}
    \vec{P} \sim \vec{\phi} &\dist g_P(\vec{\phi}) \\
    Y_{i,t} &\dist \BB(n_{i,t}, P_{i,t})
\end{align}
where $g_P$ is analogous to $g_Z$ in providing a joint prior for the prevalence in each strata and day combination.

@@

The chosen approach should have the following properties.
\begin{enumerate}
    \item The smoothing model over time should penalise changes from continuous exponential growth.
        Penalising changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
        Furthermore, such penalisation has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
    \item Limit estimates to epidemiologically and biologically plausible values.
        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
        % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias.
    \item Borrow information across the different covariates.
        The borrowing should penalise changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
        A constant distribution is predicted by the theoretical study of epidemics, for example the dynamics of stratified, mechanistic transmission models (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) over short periods of time with negligible behavioural and immunity changes.
    \item Be computationally efficient
        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
\end{enumerate}

Properties 1 and 2 suggest a model of the form $\logit(\pi_{i,t}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalised (\ie changes away from linear are penalised).
The $\logit$ link function both limits $\pi_{i,t}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{i,t}$~\autocite{ealesAppropriately}.
Combined with property 4, the form of $g$ the null model is $g(i, t) = a_0 + a_t t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified; that is, changes away from this model should be penalised.

@@

In practice, this means that $\eta_j$ is a one-to-one transformation of the population prevalence in stratum $i$ on day $t$.
If $\eta_j = \logit(\pi_{i,t})$, where $\pi$ is prevalence, then a model of $\eta_j = g(i, t)$, as before, is a LGM.
However, if $\pi$ is instead representing incidence proportion, then no such linear relationship between $\eta_j$ and $g$ exists.
In particular, if $\eta_j$ is the prevalence, then:
\begin{align}
    \eta_{i,t}
    &= \sum_{k=-\infty}^0 \pi_{ik} S(k) \\
    &= \sum_{k=-\infty}^0 \expit(g(i, k)) S(k).
\end{align}
As $g(i, k)$ is linear in the model parameters, $\vec{z}$, then $\eta_{i,t}$ is not, violating the requirements of a LGM.

@@

First, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
I solve this system using forward substitution, calculating the value of $Z_t$ in \cref{E-inc-prev:eq:forward-substitute} sequentially for $t = 1, \dots, T$.
\Cref{E-inc-prev:eq:forward-substitute} is repeated here for convenience:
\begin{align}
Z_t
&= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
&= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
\label{transmission:eq:forward-substitute}
\end{align}
% \Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
Dividing each side by the total number of individuals in each stratum transforms number of prevalent individuals to prevalence and incidence to incidence proportion.

Violation of assumptions in various parts of this procedure can lead to the second term in \cref{transmission:eq:forward-substitute} to be greater than the first, giving negative incidence.
In this case, the incidence prior to day $t$ imply that the prevalence on day $t$ is higher than that predicted by the fitted regression model for this strata/day combination.
There are two sets of assumptions which are most likely to be violated.
The first set of assumptions is in constructing survival matrix, $\matr{S}$, which is assumed to be known with no uncertainty in this procedure.
However, \cref{E-imperf-test} showed that the estimate of $S(t)$ is sensitive to assumptions made, such as the assumed $\psens$ and strength of priors used.
The second set of assumptions are in specifying the statistical model used to estimate prevalence, for example the form of the smoothing used.
In this context, that means the prior $\pi_P$ places non-zero mass on prevalence values which imply negative incidence under the other model assumptions.

@@

Once the incidence proportion in each stratum has been calculated using this procedure, I poststratify it to the CIS target population and subpopulations of interest.
This is done using the methodology explained in \cref{E-biology-data:sec:MRP}.

@@

Various alternative inference techniques could be considered.
If the infection process is considered a Poisson process (defined in \cref{E-inc-prev:sec:poisson-process}), then, conditional on the process's intensity, there is independence between individuals' infection status.
The conditional independence implies the incidence proportion and prevalence are the mean of many independent Bernoulli variables.
Therefore, the Central Limit Theorem implies they are approximately normally distributed.
Exploiting this property could enable efficient approximations to the posterior.
For example, a Laplace approximation or variational inference.
Further work, to check the quality of these approximations, is required.

@@

In this section, I apply the general mechanistic transmission modelling framework from \cref{E-SEIR} to the CIS prevalence data.
The application of this framework follows \textcite{birrellRealtime}.
The largest modification from the generic framework presented in \cref{E-SEIR} is the parameterisation of the WAIFW matrix (\cref{SEIR:sec:WAIFW-application}).
A novel approach (\cref{SEIR:sec:observation}) links this transmission model to the CIS prevalence data; the approach is generalisable to other studies collecting representative prevalence data, \eg REACT.
In the mechanistic modelling literature, this part of the model is referred to as the observation model, however, in the terminology of \cref{E-inc-prev}, it includes both the prevalence process and the observation process.
Having specified the full model (\cref{SEIR:sec:methods-application}), I conducted a simulation study to validate the inference process and identifiability (\cref{SEIR:sec:sim-study}), before applying the model to the CIS data (\cref{SEIR:sec:application}).
In \cref{SEIR:sec:discussion}, I discuss the results; in \cref{transmission:sec:comparison} I compare the results to those in the previous section.

@@

I model each region fully independently, \ie with no shared parameters or transmission between regions.
As well as the motivation given in \cref{E-backcalc}, independent regions allows heterogeneity in the parameters between regions, which could arise from a range of socioeconomic factors.
Furthermore, the assumption of homogeneous mixing within age groups is more plausible for regions than the whole of England.
By modelling each region independently, I assume that the epidemics in each region do not interact.
Clearly, this is not realistic as travel between regions of England is common and unrestricted.
However, except in an epidemic's early stages, the interaction between large regions has a negligible impact on the epidemics progression~\autocite[e.g.][]{birrellRealtimea,gogSpatial,eggoSpatial}.

@@

A parsimonious model is needed to parameterise the WAIFW matrix.
In \cref{E-SEIR:sec:structured-populations}, I explained how the elements of the WAIFW matrix, $\matr{\beta}$ can be decomposed as $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'} / N_{a}$ where\todo{should I repeat these definitions? they already exist in ch 3}: $\kappa_{aa'}$ is the mean rate at which an individual in stratum $a'$ makes contact with individuals in stratum $a$; $q_{aa'}$ is the probability of transmission upon contact between an infectious individual in stratum $a'$ and a susceptible individual in stratum $a$; and $N_{a}$ is the total number of individuals in stratum $a$.
For many epidemics, where the population behaviour is fairly constant over time, a constant $\matr{\beta}$ is a good approximation.
However, in SARS-CoV-2, population behaviour changed dramatically, both spontaneously (\eg due to fear) and in response to government-imposed restrictions.
Therefore, following \textcite{birrellRealtime}, I introduce a time-dependence giving: $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'}(t) N_{a'} / N_a$; $N_a$ and $N_{a'}$ are constant because the change in the total population is negligible over the length of time considered here.

@@

% Here, I extend the system of ODEs to include compartments representing RT-PCR positive individuals.
% The proportion of RT-PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are RT-PCR positive as the mean of the Beta distribution.
One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
Here, the daily incidence proportion for each stratum derived by solving the system of ODEs governing the transmission model (see \cref{E-SEIR:sec:SIR}), and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
However, this approach is computationally expensive, likely prohibitively so if performed within a MCMC loop.

A computationally cheaper option, which I will take, is to approximate this convolution by adding compartments to the ODEs.
This approach allows an infected, pre-positive compartment, which I denote $p_0$, to be included.
Up until now, this stage of the disease (shown in \cref{E-biology-data:fig:natural-history}) has been neglected, assumed to be of negligble duration.
An advantage of this modelling approach is that it can be included without complications (\eg issues of numerical stability).

@@

I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{E-biology-data:sec:clustering}).
If on each day $t$, each strata $i$ is observed to have $y_{at}$ positives out of $n_{at}$ tests, then the likelihood is:
\begin{align}
    \prod_{a,t} \BB (y_{at} \mid n_{at}, \bar{p}_{at}, \rho)
    \label{SEIR:eq:likelihood}
\end{align}
where: $\bar{p}_{at}$ is the mean proportion of individuals in strata $a$ that are RT-PCR positive on day $t$; and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
Denoting the proportion of stratum $a$ in state $p_k$ as $p_{ka}(\tau)$, then $\bar{p}_{at} = \int_{t}^{t+1} \sum_{k=1}^l p_{ka}(\tau) d\tau$.

@@

The remaining details of the transmission model structure follow \textcite{birrellRealtime}.
The transmission model is a SEIR model with two latent and two infectious states.
To form whole of England estimates, each region is modelled independently with the results aggregated, using the poststratification methodology explained in \cref{E-backcalc}.

@@

Priors for the parameters are given in \cref{SEIR:table:priors}.
The parameters governing the duration of positivity ($\omega$s) were explained previously (see \cref{SEIR:table:ec-params}).
The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors or fixed values are used for them.
$\sigma_\epsilon$ and $\rho$ are given priors that favour a simpler model.
The prior on $\sigma_\epsilon$ is a penalised complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
The prior on $\rho$ favours a null model of a binomial likelihood.
The remaining parameters have weakly informative priors.

@@

\begin{landscape}
\begin{table}
\begin{tabular}{l c l l}
    Parameter description & Symbol & Distribution & Comment \\
    \hline \\
    Mean latent period & $1/\sigma$ & 3.5 days (fixed) & Based on \textcite{zhaoEstimating} \\
    Mean infectious period & $1/\gamma$ & 4 days (fixed) & Based on \textcite{zhaoEstimating} \\
    Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity-prior} & \\
    Initial proportion infectious & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
    Susceptibility of children & $q_c$ & $\LN(-0.4325, 0.1174)$ & Based on \textcite{vinerTransmission}  \\
    Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
    Standard deviation of random walk steps on log scale & $\sigma_\epsilon$ & $\Exponential(80)$ & Penalised complexity prior \\
    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & Penalising deviations from binomial
\end{tabular}
\caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterisations see \cref{E-distributions}.}
\label{SEIR:table:priors}
\end{table}
\begin{table}

@@

I compute the posterior distribution using MCMC.
Specifically, I use an adaptive random-walk Metropolis-Hastings algorithm (discussed in \cref{E-MCMC}).
The proposal distribution is a multivariate normal that adapts to the covariance of the posterior.
The adaption algorithm is based on \textcite[algorithm 4]{andrieuTutorial}, with the implementation based on \textcite{ghoshApproximate}'s work.
Proposals are accepted or rejected using the standard Metropolis-Hastings (MH) acceptance probability.
See \cref{SEIR:MCMC-algorithm} for full details.

>>>>

In this chapter, I use the estimate of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
I will show that the CIS data is sufficient, demonstrating the value of surveys of this type for pandemic surveillance.

@@

First, in \cref{backcalc}, I use a back-calculation approach, based on \cref{E-inc-prev:eq:EPt}.
This approach enables estimation of incidence within (sub)populations of England.
There are several attractive features about this approach in the context here.
The backcalculation approach makes no assumptions about the transmission process; the assumptions have the potential to introduce bias.
It can also easily incorporate arbitrary covariates to adjust for non-response bias.
The scale of the CIS and desire to correct for non-response bias presents challenges for this methodology, especially within the limited computation of the SRS environment.

@@

The first approach, which I call the \emph{incidence-inferred} approach, is a single statistical model where incidence is inferred, with prevalence as a derived quantity.
The basis of the statistical model, in a Bayesian framework, would be as follows.
\begin{align}
    \vec{Z} \mid \vec{\phi} &\dist g_Z(\vec{\phi}) \\
    P_{i,t} &= \matr{S} \vec{Z}_i \\
    Y_{i,t} \mid \vec{P} &\dist \BB(n_{i,t}, P_{i,t})
\end{align}
where $\vec{Z}$ is a vector of incidence proportion for each strata and day defined by some model $g$ with parameters $\vec{\phi}$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\matr{S}$ is the survival matrix defined in \cref{E-inc-prev:eq:Smatrix}; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
The model $g_Z$ would be specified to share information across time and/or strata.
This both helps stabilise the estimation (see \cref{intro:sec:disease-model}) and allows borrowing of information across strata.
Borrowing information is required because some strata and time combinations have few positives (fewer than five per day).
The relationship between $P$ and $Z$ is from \cref{E-inc-prev:eq:EPt-matrix}, although dividing both sides by $\Npop$ to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.

The second approach, which I call the \emph{prevalence-inferred} approach takes a two-step approach.
First, it uses the statistical model from \cref{E-biology-data:sec:MRP} to estimate prevalence.
Then, it transforms the posterior distribution of prevalence to a posterior distribution of incidence using the forward substitution as in \cref{E-inc-prev:eq:forward-substitute}.
At a high-level, the statistical model is:
\begin{align}
    \vec{P} \sim \vec{\phi} &\dist g_P(\vec{\phi}) \\
    Y_{i,t} &\dist \BB(n_{i,t}, P_{i,t})
\end{align}
where $g_P$ is analogous to $g_Z$ in providing a joint prior for the prevalence in each stratum and day combination.

@@

The chosen approach should have the following properties.
\begin{enumerate}
    \item The smoothing model over time should penalize changes from continuous exponential growth.
        penalizing changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
        Furthermore, such penalization has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
    \item Limit estimates to epidemiologically and biologically plausible values.
        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
        % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias.
    \item Borrow information across the different covariates.
        The borrowing should penalize changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
        A constant distribution is predicted by the theoretical study of epidemics, for example the dynamics of stratified, mechanistic transmission models (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) over short periods of time with negligible behavioural and immunity changes.
    \item Be computationally efficient.
        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
\end{enumerate}

Properties 1 and 2 suggest a model of the form $\logit(\pi_{i,t}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalized (\ie changes away from linear are penalized).
The $\logit$ link function both limits $\pi_{i,t}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{i,t}$~\autocite{ealesAppropriately}.
Combined with property 4, the form of $g$ the null model is $g(i, t) = a_0 + a_t t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified; that is, changes away from this model should be penalized.

@@

In practice, this means that $\eta_j$ is a one-to-one transformation of the population prevalence in stratum $i$ on day $t$.
If $\eta_j = \logit(\pi_{i,t})$, where $\pi$ is prevalence, then a model of $\eta_j = g(i, t)$, as before, is an LGM.
However, if $\pi$ is instead representing incidence proportion, then no such linear relationship between $\eta_j$ and $g$ exists.
In particular, if $\eta_j$ is the prevalence, then:
\begin{align}
    \eta_{i,t}
    &= \sum_{k=-\infty}^0 \pi_{ik} S(k) \\
    &= \sum_{k=-\infty}^0 \expit(g(i, k)) S(k).
\end{align}
As $g(i, k)$ is linear in the model parameters, $\vec{z}$, then $\eta_{i,t}$ is not, violating the requirements of an LGM.

@@

First, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
I solve this system using forward substitution, calculating the value of $Z_t$ in \cref{E-inc-prev:eq:forward-substitute} sequentially for $t = 1, \dots, T$.
\Cref{E-inc-prev:eq:forward-substitute} is repeated here for convenience:
\begin{align}
Z_t
&= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
&= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
\label{transmission:eq:forward-substitute}
\end{align}
% \Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
Dividing each side by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.

Violation of assumptions in various parts of this procedure can lead to the second term in \cref{transmission:eq:forward-substitute} to be greater than the first, giving negative incidence.
In this case, the incidence prior to day $t$ imply that the prevalence on day $t$ is higher than that predicted by the fitted regression model for this strata/day combination.
There are two sets of assumptions which are most likely to be violated.
The first set of assumptions is in constructing survival matrix, $\matr{S}$, which is assumed to be known with no uncertainty in this procedure.
However, \cref{E-imperf-test} showed that the estimate of $S(t)$ is sensitive to assumptions made, such as the assumed $\psens$ and strength of priors used.
The second set of assumptions are in specifying the statistical model used to estimate prevalence, for example the form of the smoothing used.
In particular, the prior for $\pi_P$ places non-zero mass on prevalence values which imply negative incidence under the other model assumptions.

@@

Once the incidence proportion in each stratum have been calculated using this procedure, I poststratify it to the CIS target population and subpopulations of interest.
This is done using the methodology explained in \cref{E-biology-data:sec:MRP}.

@@

Various alternative inference techniques could be considered.
If the infection process is considered a Poisson process (defined in \cref{E-inc-prev:sec:poisson-process}), then, conditional on the process's intensity, there is independence between individuals' infection status.
The conditional independence implies the incidence proportion and prevalence are the mean of many independent Bernoulli variables.
Therefore, the Central Limit Theorem implies they are approximately normally distributed.
Exploiting this property could enable efficient approximations to the posterior.
For example, a Laplace approximation or variational inference.
Further work to check the quality of these approximations, is required.

@@

In this section, I apply the general mechanistic transmission modelling framework from \cref{E-SEIR} to the CIS prevalence data.
The application of this framework follows \textcite{birrellRealtime}.
The largest modification from the generic framework presented in \cref{E-SEIR} is the parameterization of the WAIFW matrix (\cref{SEIR:sec:WAIFW-application}).
A novel approach (\cref{SEIR:sec:observation}) links this transmission model to the CIS prevalence data; the approach is generalisable to other studies collecting representative prevalence data, \eg REACT.
In the mechanistic modelling literature, this part of the model is referred to as the observation model, however, in the terminology of \cref{E-inc-prev}, it includes both the prevalence process and the observation process.
Having specified the full model (\cref{SEIR:sec:methods-application}), I conducted a simulation study to validate the inference process and identifiability (\cref{SEIR:sec:sim-study}), before applying the model to the CIS data (\cref{SEIR:sec:application}).
In \cref{SEIR:sec:discussion}, I discuss the results; in \cref{transmission:sec:comparison} I compare the results to those in the previous section.

@@

I model each region fully independently, \ie with no shared parameters or transmission between regions.
As well as the motivation given in \cref{E-backcalc}, independent regions allow heterogeneity in the parameters between regions, which could arise from a range of socioeconomic factors.
Furthermore, the assumption of homogeneous mixing within age groups is more plausible for regions than the whole of England.
By modelling each region independently, I assume that the epidemics in each region do not interact.
Clearly, this is not realistic as travel between regions of England is common and unrestricted.
However, except in an epidemic's early stages, the interaction between large regions has a negligible impact on the epidemic's progression~\autocite[e.g.][]{birrellRealtimea,gogSpatial,eggoSpatial}.

@@

A parsimonious model is needed to parameterize the WAIFW matrix.
In \cref{E-SEIR:sec:structured-populations}, I explained how the elements of the WAIFW matrix, $\matr{\beta}$ can be decomposed as $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'} / N_{a}$ where\todo{should I repeat these definitions? they already exist in ch 3}: $\kappa_{aa'}$ is the mean rate at which an individual in stratum $a'$ makes contact with individuals in stratum $a$; $q_{aa'}$ is the probability of transmission upon contact between an infectious individual in stratum $a'$ and a susceptible individual in stratum $a$; and $N_{a}$ is the total number of individuals in stratum $a$.
For many epidemics, where the population behaviour is fairly constant over time, a constant $\matr{\beta}$ is a good approximation.
However, in SARS-CoV-2, population behaviour changed dramatically, both spontaneously (\eg due to fear) and in response to government-imposed restrictions.
Therefore, following \textcite{birrellRealtime}, I introduce a time-dependence giving: $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'}(t) N_{a'} / N_a$; $N_a$ and $N_{a'}$ are constant because the change in the total population is negligible over the length of time considered here.

@@

% Here, I extend the system of ODEs to include compartments representing RT-PCR positive individuals.
% The proportion of RT-PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are RT-PCR positive as the mean of the Beta distribution.
One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
Here, the daily incidence proportion for each stratum derived by solving the system of ODEs governing the transmission model (see \cref{E-SEIR:sec:SIR}), and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
However, this approach is computationally expensive, likely prohibitively expensive if performed within a MCMC loop.

A computationally cheaper option, which I will take, is to approximate this convolution by adding compartments to the ODEs.
This approach allows an infected, pre-positive compartment, which I denote $p_0$, to be included.
Up until now, this stage of the disease (shown in \cref{E-biology-data:fig:natural-history}) has been neglected, assumed to be of negligible duration.
An advantage of this modelling approach is that it can be included without complications (\eg issues of numerical stability).

@@

I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{E-biology-data:sec:clustering}).
If on each day $t$, each stratum $i$ is observed to have $y_{at}$ positives out of $n_{at}$ tests, then the likelihood is:
\begin{align}
    \prod_{a,t} \BB (y_{at} \mid n_{at}, \bar{p}_{at}, \rho)
    \label{SEIR:eq:likelihood}
\end{align}
where: $\bar{p}_{at}$ is the mean proportion of individuals in strata $a$ that are RT-PCR positive on day $t$; and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
Denoting the proportion of stratum $a$ in state $p_k$ as $p_{ka}(\tau)$, then $\bar{p}_{at} = \int_{t}^{t+1} \sum_{k=1}^l p_{ka}(\tau) d\tau$.

@@

The remaining details of the transmission model structure follow \textcite{birrellRealtime}.
The transmission model is an SEIR model with two latent and two infectious states.
To form whole of England estimates, each region is modelled independently with the results aggregated, using the poststratification methodology explained in \cref{E-backcalc}.

@@

Priors for the parameters are given in \cref{SEIR:table:priors}.
The parameters governing the duration of positivity ($\omega$s) were explained previously (see \cref{SEIR:table:ec-params}).
The initial conditions and transition rate parameters are unidentifiable in an SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors or fixed values are used for them.
$\sigma_\epsilon$ and $\rho$ are given priors that favour a simpler model.
The prior on $\sigma_\epsilon$ is a penalized complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
The prior on $\rho$ favours a null model of a binomial likelihood.
The remaining parameters have weakly informative priors.

@@

\begin{landscape}
\begin{table}
\begin{tabular}{l c l l}
    Parameter description & Symbol & Distribution & Comment \\
    \hline \\
    Mean latent period & $1/\sigma$ & 3.5 days (fixed) & Based on \textcite{zhaoEstimating} \\
    Mean infectious period & $1/\gamma$ & 4 days (fixed) & Based on \textcite{zhaoEstimating} \\
    Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity-prior} & \\
    Initial proportion infectious & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
    Susceptibility of children & $q_c$ & $\LN(-0.4325, 0.1174)$ & Based on \textcite{vinerTransmission}  \\
    Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
    Standard deviation of random walk steps on log scale & $\sigma_\epsilon$ & $\Exponential(80)$ & penalized complexity prior \\
    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & penalizing deviations from binomial
\end{tabular}
\caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterizations see \cref{E-distributions}.}
\label{SEIR:table:priors}
\end{table}
\begin{table}

@@

I compute the posterior distribution using MCMC.
Specifically, I use an adaptive random-walk Metropolis--Hastings algorithm (discussed in \cref{E-MCMC}).
The proposal distribution is a multivariate normal that adapts to the covariance of the posterior.
The adaptive algorithm is based on \textcite[algorithm 4]{andrieuTutorial}, with the implementation based on \textcite{ghoshApproximate}'s work.
Proposals are accepted or rejected using the standard Metropolis--Hastings (MH) acceptance probability.
See \cref{SEIR:MCMC-algorithm} for full details.

\begin{algorithm}
 Set $\vec{X_0}$ to an initial value of the parameter vector \;
 $\vec{\mu_0} = \vec{X_0}$ \;
 $\matr{\Sigma_0} = \text{diag}(\vec{\mu_0})$ \;
 $\lambda_0 = 1$ \;
 \For{$i = 1, \dots, M$}{
  sample $\vec{Y_}i \sim \N(\vec{\mu_{i-1}}, \lambda_{i-1}\matr{\Sigma_{i-1}})$\;
  set $\vec{X_i}$ to $\vec{Y_i}$ or $\vec{X_{i-1}}$ using a MH acceptance step\;
  update the proportion of proposals accepted so far, $\alpha$ \;
  \eIf(\tcc{No adaptation for 200 iterations}){$i \leq 200$}{
    $\lambda_i = \lambda_{i-1}$ \;
    $\vec{\mu_i} = \vec{\mu_{i-1}}$ \;
    $\matr{\Sigma_i} = \matr{\Sigma_{i-1}}$ \;
   }{
    $\gamma_i = (i - 200)^{-0.6}$ \;
    $\log \lambda_i = \log \lambda_{i-1} + \gamma_i(\alpha - 0.234)$ \;
    $\vec{\mu_i} = (1 - \gamma_i) \vec{\mu_{i-1}} + \gamma_i \vec{X_i}$ \;
    $\matr{\Sigma_i} = (1 - \gamma_i) \matr{\Sigma_{i-1}} + \gamma_i (\vec{X_i} - \vec{\mu_i})(\vec{X_i} - \vec{\mu_i})^T$ \;
  }
 }
 \caption{Algorithm for adaptive random-walk Metropolis--Hastings. $\vec{\mu_i}$ and $\matr{\Sigma_i}$ are an estimate of the mean and covariance of the posterior distribution using information up to iteration $i$. $\text{diag}(\vec{\mu_0})$ is the diagonal matrix with diagonal entries equal to $\vec{\mu_0}$. $\lambda_i$ is the scale parameter of the proposal distribution at iteration $i$, tuned to try and ensure an optimal proportion of proposals are accepted (23.4\%). $\gamma_i$ is the learning rate, which determines how much adaptation occurs. $\gamma_i \to 0$ as $i \to \infty$ so the rate of adaptation is \emph{vanishing}. Vanishing adaptation is required for the algorithm to converge to the target distribution~\autocite[section 3]{andrieuTutorial}.}
 \label{SEIR:MCMC-algorithm}
\end{algorithm}
